### 2022-04-29 paper/localpsf.tex
wrote up spd modification technique in appendix

If $\Aop$ is symmetric positive semidefinite, we symmetrize $\mathbf{A}_H$ as follows
\begin{equation*}
\mathbf{A}_H^{\text{sym}} := \frac{1}{2}\left(\mathbf{A}_H + \mathbf{A}_H^T\right),
\end{equation*}
then we use the high order rational function spectral projection method from \cite{HackbuschKress07} to replace (or approximately replace) the negative eigenvalues of $\mathbf{A}_H^{\text{sym}}$ with zero. In detail, let $f$ be the following scalar rational function,
\begin{equation*}
	f(\lambda) := \begin{cases}
		\lambda, & \lambda \ge 0, \\
		0, & \lambda < 0,
	\end{cases}
\end{equation*}
and extend $f$ to matrices via functional calculus. The matrix $f(\mathbf{A}_H^{\text{sym}})$ is the ideal positive semidefinite modification of $\mathbf{A}_H^{\text{sym}}$, because $f(\mathbf{A}_H^{\text{sym}})$ is the same as $\mathbf{A}_H^{\text{sym}}$ except all the negative eigenvalues are replaced with zero. Computing $f(\mathbf{A}_H^{\text{sym}})$ directly is difficult, but we may approximate $f$ by the high order rational function
\begin{equation*}
	f(\lambda) \approx \widetilde{f}(\lambda) := \frac{\lambda}{1 + \left(\frac{2 \lambda - (b+a)}{b-a}\right)^{2^k} },
\end{equation*}
where $k$ is a positive integer, $a$ is either zero, or a small positive real number, and $b$ is a positive real number that is larger than the largest eigenvalue of $\mathbf{A}_H^{\text{sym}}$. If $k$ is sufficiently large, then $\widetilde{f}(\lambda) \approx \lambda$ for $\lambda \in (a,b)$ and $\widetilde{f}(\lambda) \approx 0$ for $\lambda$ outside of $[a,b]$, so $\widetilde{f}(\mathbf{A}_H^{\text{sym}})$ is a good approximation of  $f(\mathbf{A}_H^{\text{sym}})$. This is illustrated in Figure \ref{fig:spd_rat}.  In practice we use $k=7$, $a=0$ and $b=1.5~ \lambda_\text{max}$, where $\lambda_\text{max}$ is the largest eigenvalue of $\mathbf{A}_H^{\text{sym}}$, which is computed using the Lanczos iteration. 
\begin{figure}
	\centering
	\includegraphics[scale=0.5]{spd_rational_function.pdf}
	\caption{Rational function $\widetilde{f}(\lambda)$ for $M=1$ and $k=3,5,7$. As $k$ increases, $\widetilde{f}(\lambda)$ approaches the sawtooth function supported on $(0,M)$. }
	\label{fig:spd_rat}
\end{figure}

Using $\widetilde{f}$ instead of $f$ permits us to efficiently compute $\widetilde{f}(\mathbf{A}_H^{\text{sym}})$ using H-matrix methods. Replacing constants with scaled identity matrices, scalar powers and multiplication with matrix powers and multiplication, and scalar division with matrix inversion, the extension of $\widetilde{f}$ to matrices becomes
\begin{equation}
\label{eq:rational343}
\widetilde{f}(\mathbf{A}_H^{\text{sym}}) = \left(\mathbf{I} + \mathbf{X}^{2^k}\right)^{-1} \mathbf{A}_H^{\text{sym}},
\end{equation}
where
\begin{equation*}
	\mathbf{X} := \left(2 \mathbf{A}_H^{\text{sym}} - (b+a) \mathbf{I}\right)/(b-a).
\end{equation*}
We compute $\mathbf{X}$ using fast H-matrix scalar multiplication and addition. Then we form the $\mathbf{X}^{2^k}$ via repeated squaring,
\begin{align*}
	\mathbf{X}^2 &= \mathbf{X} \mathbf{X} \\
	\mathbf{X}^4 &= \mathbf{X}^2 \mathbf{X}^2 \\
	&\vdots \\
	\mathbf{X}^{2^k} &= \mathbf{X}^{2^{k-1}} \mathbf{X}^{2^{k-1}},
\end{align*}
where each of these matrix squarings is performed using fast H-matrix multiplication. Finally, $\widetilde{f}(\mathbf{A}_H^{\text{sym}})$ is formed from $\mathbf{X}^{2^k}$ and $\mathbf{A}_H^{\text{sym}}$ per \eqref{eq:rational343}, using fast H-matrix operations to perform the required matrix-matrix addition, matrix inversion, and matrix-matrix multiplication. At the end, the matrix is symmetrized once again, to correct for any small non-symmetries that are introduced due to truncation error in the H-matrix arithmetic. Once constructed, we use the symmetric positive semidefinite matrix $\widetilde{f}(\mathbf{A}_H^{\text{sym}})$ in place of $\mathbf{A}_H$ in further linear algebra operations.

>>>>

If $\Aop$ is symmetric, we symmetrize $\mathbf{A}_H$ as follows
\begin{equation*}
\mathbf{A}_H^{\text{sym}} := \frac{1}{2}\left(\mathbf{A}_H + \mathbf{A}_H^T\right).
\end{equation*}
If $\Aop$ is also positive semi-definite, we use the high order rational function spectral projection method from \cite{HackbuschKress07} to flip (or approximately flip) the negative eigenvalues of $\mathbf{A}_H^{\text{sym}}$ to be positive. In detail, let $\Pi$ be the indicator function for the interval $(a,b)$,
\begin{equation*}
	\Pi(\lambda) = \begin{cases}
		1, & \lambda \in (a,b), \\
		0, & \text{otherwise},
	\end{cases}
\end{equation*}
and extend $\Pi$ to matrices via functional calculus. The matrix $\Pi(\mathbf{A}_H^{\text{sym}})$ is the spectral projector onto the eigenspace of $\mathbf{A}_H^{\text{sym}}$ corresponding to eigenvalues of $\mathbf{A}_H^{\text{sym}}$ that reside in $(a,b)$. If we choose $a < \lambda_\text{min}$ and $b=0$, where $\lambda_\text{min}$ is the smallest eigenvalue of $\mathbf{A}_H^{\text{sym}}$, then the matrix absolute value of $\mathbf{A}_H^{\text{sym}}$ is given by
\begin{equation}
\label{eq:I_Pi_abs}
	|\mathbf{A}_H^{\text{sym}}| = \left(\mathbf{I} - 2\Pi(\mathbf{A}_H^{\text{sym}})\right)\mathbf{A}_H^{\text{sym}}.
\end{equation}
That is, $|\mathbf{A}_H^{\text{sym}}|$ is the matrix that has the same eigenvectors as $\mathbf{A}_H^{\text{sym}}$, but the eigenvalues of $|\mathbf{A}_H^{\text{sym}}|$ are the absolute values of the eigenvalues of $\mathbf{A}_H^{\text{sym}}$.

Computing $\Pi(\mathbf{A}_H^{\text{sym}})$ directly is difficult, but we may approximate $\Pi$ by the following high order rational function:
\begin{equation*}
	\Pi(\lambda) \approx \Pi_k(\lambda) := \frac{1}{1 + T(\lambda)^{2^k} }, \qquad \text{where} \qquad T(\lambda) := \frac{2 \lambda - (b+a)}{b-a}
\end{equation*}
and $k$ is a positive integer; the larger $k$ is, the better the approximation $\Pi \approx \Pi_k$. This is illustrated in Figure \ref{fig:spd_rat}. For $k \ge 1$, $\Pi_k(\lambda)> 1/2$ for $\lambda \in (a,b)$, so choosing $a < \lambda_\text{min}$, $b=0$, and using $\Pi_k$ instead of $\Pi$ in \eqref{eq:I_Pi_abs} yields a matrix that is positive semidefinite. Replacing constants with scaled identity matrices, scalar powers and multiplication with matrix powers and multiplication, and scalar division with matrix inversion, the extension of $\Pi_k$ to matrices becomes
\begin{equation}
\label{eq:rational343}
\Pi_k\left(\mathbf{A}_H^{\text{sym}}\right) = \left(\mathbf{I} + \mathbf{T}^{2^k}\right)^{-1},
\end{equation}
where
\begin{equation*}
\mathbf{T} := \left(2 \mathbf{A}_H^{\text{sym}} - (b+a) \mathbf{I}\right)/(b-a).
\end{equation*}

\begin{figure}
	\centering
	\includegraphics[scale=0.5]{spd_rational_function.pdf}
	\caption{Rational function $\Pi_k(\lambda)$ for $a=-1$, $b=0$, and $k \in \{1,2,3\}$. As $k$ increases, $\Pi_k$ approaches the indicator function for the set $(a,b)$. }
	\label{fig:spd_rat}
\end{figure}

Using $\Pi_k$ instead of $\Pi$ permits us to efficiently approximate $|\mathbf{A}_H^{\text{sym}}|$ using H-matrix methods. 
First, we compute $\lambda_\text{min}$ using the Lanczos method, and set $a=\gamma \lambda_\text{min}$ and $b=0$, where $\gamma\ge 1$ is a parameter that must be chosen. We use $\gamma=1.5$, though the method is relatively insensitive to $\gamma$ and works well for any $\gamma \in [1,2]$. 
Next, we compute $\mathbf{T}$ using fast H-matrix scalar multiplication and addition. We form the $\mathbf{T}^{2^k}$ via repeated squaring, $\mathbf{T}^2 = \mathbf{T} \mathbf{T}$, $\mathbf{T}^4 = \mathbf{T}^2 \mathbf{T}^2$, $\mathbf{T}^8 = \mathbf{T}^4 \mathbf{T}^4$ and so on, where each of these matrix squarings is performed using fast H-matrix multiplication. We do not need $\Pi_k$ to be an accurate approximation of $\Pi$ since the negative eigenvalues of $\mathbf{A}_H^{\text{sym}}$ are erroneous, and $\Pi_k(\lambda)$ rapidly decays to zero for positive $\lambda$, even if $k$ is small. Hence, we recommend using $k=1$ or $2$ (we use $k=1$ in our numerical results). Larger $k$ is unnecessary, and may make the method unstable. The approximate spectral projector $\Pi_k(\mathbf{A}_H^{\text{sym}})$ is formed from $\mathbf{T}^{2^k}$ per \eqref{eq:rational343}, using fast H-matrix operations to perform the required matrix-matrix addition and matrix inversion. Finally, a positive semi-definite approximation $\mathbf{A}_H^{\text{sym}+}$ of $\mathbf{A}_H^{\text{sym}}$ is formed as
\begin{equation*}
\mathbf{A}_H^{\text{sym}+} := \left(\mathbf{I} - 2\Pi(\mathbf{A}_H^{\text{sym}})\right)\mathbf{A}_H^{\text{sym}},
\end{equation*}
where fast H-matrix operations are used to perform the necessary matrix-matrix subtraction and multiplication. At the end, $\mathbf{A}_H^{\text{sym}+}$ is symmetrized once again, to correct for any small non-symmetries that are introduced due to truncation error in the H-matrix arithmetic. Once constructed, we use the symmetric positive semidefinite matrix $\mathbf{A}_H^{\text{sym}+}$ in place of $\mathbf{A}_H$ in further linear algebra operations.
